\texttt{nuance}: Detection of planetary transits in the presence of correlated noise

### 2023-12-06 latex/paper.tex
feat: exocomets

In this paper, we explore the light curve characteristics for which a full-fledge transit search is necessary, and we present \nuancemethod{}\footnote{Throughout the paper, \nuancemethod{} written in italics refers to the algorithmic and analytical method, while \nuancecode{} in sans-serif refers to its implementation.}, a method to search for transit signals while modeling correlated noises in a tractable way. In \autoref{issues}, we describe the effect of correlated noise on transit light curves and the effect of its detrending on transit signals detectability. In \autoref{nuance}, we present \nuance{}, and the two main steps on which this method is based: the linear search and the periodic search. In \autoref{results}, we test the performance of \nuance{} on a wide variety of cases, and compare it to commonly used transit search algorithms. This include transits injected in synthetic datasets, but also in the TESS light curves of 438 rapidly-rotating M dwarfs. Finally, in \autoref{discussion}, we discuss the results and the limitations of \nuance{}, and we conclude in .

@@

\subsection{The choice of kernel}
While not being discussed in our study, \textbf{the efficiency of \nuancemethod{} to detect transits in correlated noise is highly dependant on the design of its GP kernel}. In the ensemble comparison of \autoref{real}, the goal was to choose a kernel and an optimization strategy suited to most of the studied light curves, leading to few outliers in the results that were indicative of a badly designed and/or optimized kernel. An alternative, recommended for more realistic blind searches, is to perform model comparison on well selected kernels, and to adapt the optimization strategy to each dataset.\\\\
When using \nuancecode{} on TESS light curves for example, it must be noted that the observed light curve variability might originate from a contaminated photometric aperture, so that a physically-interpretable GP kernel representing a single star activity is not necessarily appropriate. On the other hand, a single squared exponential GP kernel might also be sufficient for some applications.

\subsection{Prospects}
The present implementation of \nuancemethod{} has the potential to be extended to be used beyond the search of periodic transits. Here are ideas of possible use and extensions, from the most straightforward to the most ambitious:
\begin{enumerate}
    \item In order to compare \nuancecode{} to BLS-based methods, we injected and retrieved only box-shaped transits. However, similarly to the Transit-Least-Square algorithm from \cite{tls}, \textbf{limb-darkened transits can serve as base model} in the linear search and are expected to improve the transit search in the same way TLS provided an improvement over BLS.
    \item The linear search of \nuancemethod{} is a single-event detection algorithm that can be used to search for single transit events, but also \textbf{detect transiting exocomets and flares}, by simply adapting the base astrophysical model in the last colum of the design matrix $\bf{X}$. While not being tested and benchmarked, the \nuancecode{} implementation already integrates this feature (make a tutorial!).
    \item During the periodic search, no prior about the transit duration related to the orbital period of the planet was used. This was done in order to allow the detection of grazing transiting exoplanets that would produce shorter timescale transits compared to what is expected from a circular orbit. However, adding such prior, \textbf{using the distribution of known exoplanets eccentricities}, might produce less spurious periodogram peaks and be very beneficial for the automatic search of transits in large datasets. Another idea, similar to the one employed in \cite{foreman2016}, is to leverage \textbf{model comparison in order to reject single transits that are better described by the GP model alone}. Both ideas come at no cost given our modeling approach.
    \item Finally, the formalism of \nuancemethod{} could be adapted and used to \textbf{search for exoplanets featuring transit time variations (TTV)}. Indeed, this application only requires a modification of the periodic search, as maximum likelihood peaks close to linearly predicted transit epochs may be considered. This could be done either with a special nearest-neighbor algorithm or with a convolution of the computed likelihood grid with a Gaussian kernel. To maximize efficiency and interpretability, we would recommend these approaches to be explored analytically, rather than using a data-driven treatment of the linear search products.
\end{enumerate}

\section{Conclusion}
This paper presents \nuancemethod{}, an algorithm designed to detect planetary transits in light curves featuring correlated noise in the form of instrumental signals and stellar variability. In this context, a conventional approach involves detrending a light curve before searching for transits using a Box-Least-Square algorithm. However, we show that the incomplete modeling of a light curve for its detrending, one that ignores the presence of transits, degrades their signal-to-noise-ratio down to the point of not being detectable. Adopting a widely used detrending stratgey, we explore the extend of this degradation on simulated light curves, and its dependance on the stellar variability characteristics.\\\\
The effectiveness of \nuancemethod{} is tested using a synthetic dataset and further validated on real TESS light curves of 438 M dwarfs. These injection-recovery tests reveal that \nuancemethod{} consistently outperforms conventional transit search techniques, especially for transits with duration exceeding one-fifth of the stellar variability timescale. In these cases, \nuancemethod{} not only identifies a higher number of true positive detections but also minimizes false positives, demonstrating its robustness and reliability.\\\\
To make \nuancemethod{} accessible for wider use, we make \nuancemethod{} publicly available through the \nuancecode{} open-source Python package. This package leverages the power of JAX, allowing for efficient computations and compatibility with distributed computing environments and GPU devices.\\\\
We acknowledge the limitations of \nuancemethod{} and its increased computational cost. In this regard, \nuancemethod{} should be used as an alternative to more traditional techniques only in the presence of substantial correlated noise. As guidelines for choosing \nuancemethod{} over other techniques are highly dependant on the type of detrending algorithm employed, we reserve this study to a future work.\\\\
Finally, we suggest future improvements and extensions of the algorithm, including its application for detecting single transits, exocomets, flares, and transit time variations (TTV), underscoring its versatility and potential for broader impact in astronomical research.
\\\\
\vfill{}
We would like to thank Julien De Wit, Prajwal Niraula and Francisco J. Pozuelos for meaningful discussions at the beginning of this project, Germain Garcia for his useful insights about numerical optimization, Michaël Gillon for his overall support, and the member of the Astronomical Data Group at the Center for Computational Astrophysics for many enriching discussions and feedback.x

>>>>

In this paper, we explore the light curve characteristics for which a full-fledge transit search is necessary, and we present \nuancemethod{}\footnote{Throughout the paper, \nuancemethod{} written in italics refers to the algorithmic and analytical method, while \nuancecode{} in sans-serif refers to its implementation.}, a method to search for transit signals while modeling correlated noises in a tractable way. In \autoref{issues}, we describe the effect of correlated noise on transit light curves and the effect of its detrending on transit signals detectability. In \autoref{nuance}, we present \nuance{}, and the two main steps on which this method is based: the linear search and the periodic search. In \autoref{results}, we test the performance of \nuance{} on a wide variety of cases, and compare our method to commonly used transit search algorithms. This include transits injected in synthetic datasets, but also in the TESS light curves of 438 rapidly-rotating M dwarfs. Finally, in \autoref{discussion}, we discuss the results and the limitations of \nuance{}, and we conclude in .

@@

\subsection{The choice of kernel and model}
While not being discussed in our study, \textbf{the efficiency of \nuancemethod{} to detect transits in correlated noise is highly dependant on the design of its GP kernel}. In the ensemble comparison of \autoref{real}, the goal was to choose a kernel and an optimization strategy suited to most of the studied light curves, leading to few outliers in the results that were indicative of a badly designed and/or optimized kernel. An alternative, recommended for more realistic blind searches, is to perform model comparison on well selected kernels, and to adapt the optimization strategy to each dataset.\\\\
When using \nuancecode{} on TESS light curves for example, it must be noted that the observed light curve variability might originate from a contaminated photometric aperture, so that a physically-interpretable GP kernel representing a single star activity is not necessarily appropriate. On the other hand, a single squared exponential GP kernel might also be sufficient for some applications.\\\\
Something to emphasize is that \nuancemethod{} cannot be used to produce reliable detrended light curves. Most of the time, the GP is flexible enough to partially model transits and can never be used to model the data without them. The idea behind \nuancemethod{} is to compute the likelihood of data against a model containing both transits and correlated noise, without ever trying to disentangle both signals. In practice, it means that it can be very hard to actually verify the presence of transits found by \nuancemethod{} visually, so that transits may be detected but remain hidden in correlated noise. This is particularly true for stars displaying very high-frequency photometric pulsations.

\subsection{Prospects}
The present implementation of \nuancemethod{} has the potential to be extended to be used beyond the search of periodic transits. Here are ideas of possible use and extensions, from the most straightforward to the most ambitious:
\begin{enumerate}
    \item In order to compare \nuancecode{} to BLS-based methods, we injected and retrieved only box-shaped transits. However, similarly to the Transit-Least-Square algorithm from \cite{tls}, \textbf{limb-darkened transits can serve as base model} in the linear search and are expected to improve the transit search in the same way TLS provided an improvement over BLS.
    \item The linear search of \nuancemethod{} is a single-event detection algorithm that can be used to search for single transit events, but also \textbf{detect transiting exocomets and flares}, by simply adapting the base astrophysical model in the last colum of the design matrix $\bf{X}$. While not being tested and benchmarked, the \nuancecode{} implementation already integrates this feature and we show an example of a search of the known exocomets of Beta Pictoris in \autoref{fig:exocomets}.
    \item During the periodic search, no prior about the transit duration related to the orbital period of the planet was used. This was done in order to allow the detection of grazing transiting exoplanets that would produce shorter timescale transits compared to what is expected from a circular orbit with a null impact parameter. However, adding such priors might produce less spurious periodogram peaks and be very beneficial for the automatic search of transits in large datasets. Another idea, similar to the one employed in \cite{foreman2016}, is to leverage \textbf{model comparison in order to reject single transits that are better described by the GP model alone}. Both ideas come at no cost given our modeling approach.
    \item Finally, the formalism of \nuancemethod{} could be adapted and used to \textbf{search for exoplanets featuring transit time variations (TTV)}. Indeed, this application only requires a modification of the periodic search, as maximum likelihood peaks close to linearly predicted transit epochs may be considered. This could be done either with a special nearest-neighbor algorithm or with a convolution of the computed likelihood grid with a Gaussian kernel. To maximize efficiency and interpretability, we would recommend these approaches to be explored analytically, rather than using a data-driven treatment of the linear search products.
\end{enumerate}

\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=\linewidth]{exocomet.pdf}
        \caption{Demonstration of the search of transiting exocomets in the TESS light curves of the star $\beta$ Pictoris (2 minutes PDCSAP data from sector 33). This star is known to display rapid $\delta$ Scuti type photometric variations with a period of about 30 minutes \citep{Lecavelier2022}. Here, we simply used the linear search of \nuancecode{} with a different base model, one that mimics the shape of transiting exocomets, to compute the SNR time-series of the signal over one sector. The maximum-SNR events are displayed in the bottom of the figure and match with the ones found by \cite{Lecavelier2022}. The GP kernel choice and its hyperparameters optimization are similar to the one presented in \autoref{real}.}
        \end{centering}
        \label{fig:exocomets} 
\end{figure}

\section{Conclusion}
This paper presents \nuancemethod{}, an algorithm designed to detect planetary transits in light curves featuring correlated noise in the form of instrumental signals and stellar variability. In this context, a conventional approach involves detrending a light curve before searching for transits using a Box-Least-Square algorithm. However, we show that the incomplete modeling of a light curve for its detrending, one that ignores the presence of transits, degrades their signal-to-noise-ratio down to the point of not being detectable. Adopting a widely used detrending stratgey, we explore the extend of this degradation on simulated light curves, and its dependance on the stellar variability characteristics, showing the need for a full-fledged transit search method like \nuancemethod{}.\\\\
The effectiveness of \nuancemethod{} is tested using a synthetic dataset and further validated on real TESS light curves of 438 rapidly-rotating M dwarfs. These injection-recovery tests reveal that \nuancemethod{} consistently outperforms conventional transit search techniques, especially for transits with duration exceeding one-fifth of the stellar variability timescale. In these cases, \nuancemethod{} not only identifies a higher number of true positive detections but also minimizes false positives, demonstrating its robustness and reliability.\\\\
We make \nuancemethod{} publicly available through the \nuancecode{} open-source Python package. This package leverages the power of JAX, allowing for efficient computations and compatibility with distributed computing environments and GPU devices. Nonetheless we acknowledge the limitations of \nuancemethod{} and its increased computational cost compared to more tr. In this regard, \nuancemethod{} should be used as an alternative to more traditional techniques only in the presence of substantial correlated noise. As guidelines for choosing \nuancemethod{} over other techniques are highly dependant on the type of detrending algorithm employed, we reserve this study to a future work.\\\\
Finally, we suggest future improvements and extensions of the algorithm, including its application for detecting single transits, exocomets, flares, and transit time variations (TTV), underscoring its versatility and potential for broader impact in astronomical research.
\\\\
\vfill{}
We would like to thank Julien De Wit, Prajwal Niraula and Francisco J. Pozuelos for meaningful discussions at the beginning of this project, Germain Garcia for his useful insights about numerical optimization, Michaël Gillon for his overall support, and the member of the Astronomical Data Group at the Center for Computational Astrophysics for many enriching discussions and feedback.x
